What We Eval in the Shadows

### 2021-04-09 paper/main.tex
background and related

This section provides a short introduction to R as it has a few
surprising features that impact the use of reflective features of the
language. We then look at the semantics of \eval in R and discuss some
design choices. Lastly, we put this paper in context of previous work.

@@

\citet{ecoop12} gave a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with rich
dynamic features that allow layering several object systems. Most data types are
sequences of primitive values constructed by calling the combination function
\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
such as \c{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by user-defined attributes, key-value pairs. For instance,
the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
call \c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
attribute ensure that the vector \c x will be treated as a $2 \times 2$ matrix
by all arithmetic operations. The \c{class} attribute is used to give a 'class',
in the object-oriented sense, to values. So, \c{class(x)<-"human"} sets the
class of \c{x} to \c{human}; classes are used for method dispatch. Every
linguistic construct is desugared to a function call, even control flow
statements, assignments, and bracketing. All functions can be shadowed or
redefined, making the language at the same time remarkably flexible and
exceeding challenging to compile statically as vividly detailed
by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
to functions. Each argument is a thunk composed of an expression, its
environment, and a slot for the result, these are called \emph{promises}. To get
the value of an argument, the corresponding promise must be forced, once forced
the promise's result is cached for future use.

@@

While a facility to transform data into code is available in many languages,
some design choices affect the expressive power of \eval and related functions.
The key design choices that we review here are the input format, the environment
in which \eval-generated code is evaluated and the reflective operations that
are available to that code. Fig.~\ref{comp} summarizes design choices for
several popular languages.

\begin{figure}[!t]\center\small
\begin{tabular}{r@{~}l|l|l|l}\hline
\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia&\cite{julia}     & expression& toplevel         & data\\
\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
\bf R&\cite{R96}  & expression& programmatic      & data, stack, envir.\\\hline
\end{tabular}
\caption{Design space of \eval}\label{comp}
\end{figure}

The input to \eval can be any format that can be turned into code. Both Julia
and R mandate that the input format be parsed expressions. JavaScript allows any
expression as a character string. Java is the most restrictive as a classloader
only accepts complete classes in the bytecode format recognized by the virtual
machine.

The environment in which \eval executes the code can be the lexically enclosing
environment of \eval (current), the global environment (toplevel), a limited
global environment (classloader), or an environment chosen by the caller of
\eval (programmatic). The most restrictive variant is that of Java, where the
scope of newly loaded code is the set of classes exposed by the current class
loader. Julia is the next most restrictive, as only global definitions can be
observed and modified by \eval. JavaScript has a variant of \eval that runs in
the current scope and a most strict mode that is restricted to global variables.
Finally, R allows the caller to choose which environment to use with \eval.

The last degree of freedom in the design of \eval is what can be done from
within the generated code. The main difference between designs lies in what part
of the system state can be accessed and modified. In R, there is very little
that cannot be observed from an \eval context and modified. The code can
override the definition of variables and functions. Also, thanks to a rich
reflective interface, code can introspect on values, walk and modify environment
chains, and access the stack frames on the call stacks and their environments.

Given the above, it is reasonable to claim that R's \eval are amongst the
languages that afford the facility the most expressive power. This comes at the
cost of the programmer's ability to reason about code in the presence of \eval.
Expressive power translates directly into limits on the usefulness of program
analysis.

By contrast, the designers of Julia chose to limit \eval. In Julia, only global
variables can be side-effected and environments cannot be readily manipulated.
This is designed to shield optimized code from some of the most pernicious uses
of the facility~\cite{oopsla18a}. Furthermore, the language provides a
versioning mechanism, called world age, to ensure that any methods defined
within an \eval only become visible at well-defined program points and thus that
recompilation does not have to occur when optimized code is
running~\cite{oopsla20a}.

@@

The R language exposes a rich reflective interface with four functions with
different semantics, \c{eval}, \c{evalq}, \c{eval.parent}, and
\c{local}. The first is the most general, taking three arguments, an expression,
its environment and an enclosing environment. The following discussion
simplifies details not relevant to this paper, the interested reader should
consult \citet{hadley}. It is defined as:

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
Parameter \c e is the expression to evaluate, \c{envir} is the evaluation
environment, and \c{encl} is used to look up variables not found in \c{envir}.
The latter is, by default, either the environment of the function that called
\eval or the top level. Parameter \c e can be of different types, for our
purposes we focus on {\tt expression}s. These can be thought of as abstract
syntax trees returned by the parser. An \c{expression} can be obtained by
calling \c{quote} with an expression or \c{parse} wit a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Even more commonly, expressions are obtained from promises using \c{substitute}.
Consider a one-parameter function \c{f}, when called with \c{a+b}, the function
returns \c{a+b} as expression.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # expression(a+b)
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise bound to variable \c x. Function \c{substitute(e,envir)} takes two
arguments, the second is used as a substitution environment for free variables
in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns a variable environment, either from the
currently executing function or the top level environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Programmers sometimes use environments as hash maps as they are the
only data type with reference semantics and have a built-in string look up.
Functions \c{parent.frame} and \c{sys.frame} return environments further up the
call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
the argument to prevent it from being evaluated in the current environment. The
\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.

@@

Richards et al.~\cite{ecoop11} provided the first large-scale study of
the runtime behavior of \eval in JavaScript. They dynamically analyzed
a corpus of the 10,000 most popular websites with an instrumented web
browser to gather execution traces. They show that \eval is pervasive
with 82\% of the most popular websites using it. The reasons for its
use include the desire to load code on demand, deserialization of JSON
dataq and lightweight meta-programming to customize web pages. While
many uses were legitimate, just as many were unnecessary and could be
replaced with equivalent and safer code. They categorized inputs to
\eval so as to cover the vast majority of input strings. Restricting
themselves to \eval in which all named variables refer to the global
scope, many patterns could be replaced by more disciplined
code~\cite{oopsla12b, moller12}. The work did not measure code
coverage, so the numbers presented are a lower bound on the possible
behaviors. Furthermore, JavaScript usage in 2011 is likely different
from today, e.g. Node.js was not covered by Richards. More details
about dynamic analysis of JavaScript can be found in~\cite{liang}.

Wang et al.~\cite{wang} analyzed use of dynamic features in 18 Python
programs to find if they affect file change-proneness. Files with
dynamic features are significantly more likely to be the subject of
changes, than other files. Chen et al. looked at the correlation
between code changes and dynamic features, including \eval, in 17
Python programs~\cite{chen}. They did not observe many uses of \eval.
Callau et al.~\cite{oscar} performed an empirical study of the usage
of dynamic features in 1,000 Smalltalk projects. While \eval itself is
not present, Smalltalk has a rich reflective interface. The authors
found that reflective are used in less than 2\% of methods. The most
common reflective method is \c{perform:}; it send a message that is
specified by a string. These features are mostly used in the core
libraries.

Bodden et al.~\cite{bodden} looked at usage of reflection in the Java
DaCapo benchmark suite. They found that dynamic loading was triggered
by the benchmark harness. The harness then executes methods via
reflection, this caused static analysis tools to generate an incorrect
call graph for the programs in DaCapo.

Morandat et al.~\cite{ecoop12} had a short section on the usage of
\eval in R. They found the it widely used in R code with 8500 call
sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call
sites account for 88\% of those. The \c{match.arg} function is the
highest used one with 54\% of all calls. In the other call sites, they
saw two uses cases. The most common is the evaluation of the source
code of a promise retrieved by \c{substitute} in a new environment;
e.g. as done in the \c{with} function. The other use case is the
invocation of a function whose name or arguments are determined
dynamically. For this purpose, R provides \c{do.call} and thus \eval
is overkill.

\todo{Look at Improving dynamic code analysis by code abstraction, VPT'21}

>>>>

This section provides a short introduction to R and the reflective features of
the language; then looks at the semantics of \eval in R and discusses design
choices; lastly, this paper is put in context of previous work.

@@

\citet{ecoop12} gave a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with a rich
complement of dynamic features expressive enough to layer several object systems
on top of the core language. Most data types are sequences of primitive values
constructed by calling the combination function \c{c}, for instance
\c{c("Ha","bye")} evaluates to a vector of two strings, constants such as \c{42}
are vectors of length one. To enable equational reasoning, values accessible
through multiple aliases are copied when written to. Furthermore, values can be
tagged by attributes, these are key-value pairs. For instance, the attribute
\c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the call
\c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
attribute turns \c x into a matrix. The \c{class} attribute gives a 'class', in
the object-oriented sense, to a value. So, \c{class(x)<-"human"} sets the class
of \c{x} to \c{human}; classes are used for method dispatch. Every linguistic
construct is desugared to a function call, even control flow statements,
assignments, and bracketing. All functions can be shadowed or redefined, making
the language at the same time remarkably flexible and exceeding challenging to
compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed
call-by-need convention for passing arguments to functions. Each argument is a
thunk composed of an expression, its environment, and a slot for the result,
these are called \emph{promises}. To get the value of an argument, the
corresponding promise must be forced, once forced the promise's result is cached
for future use.

@@

While a facility to transform data into code is available in many languages,
some design choices affect the expressive power of \eval. The key choices are
the input format, the environment in which generated code is evaluated and the
reflective operations available to that code. Fig.~\ref{comp} summarizes
several designs.

\begin{figure}[!t]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia&\cite{julia}     & expression& toplevel         & data\\
\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
\end{tabular}\caption{Design space}\label{comp}
\end{figure}

The input to \eval can be in any format convertible to code. JavaScript allows
arbitrary strings to be used. Both Julia and R are more restrictive as they
require expressions (or abstract syntax trees). Finally, Java is the most
restrictive as its classloader only accepts complete classes in bytecode form.

The choice of the environment of \eval is important as it determines how much of
a program \eval can observe as well as the reach of potential side-effects
performed by that operation. The most restrictive semantics is that of Java
where newly loaded code evaluates in the environment defined by the classes
visible from the current classloader. Julia and JavaScript's strict mode, limit
\eval to the global environment. Finally, R is the most flexible as any
accessible environment can be selected and passed to \eval.

The last degree of freedom is the expressive power of the code executed by
\eval. The main difference between languages lies in how much of the state of a
program is accessible through reflective operations. Julia, Java and JavaScript
all allow some form of introspection on the data. R is more flexible as it is
possible to inspect the program's call stack. Thus any environment in the
program can be inspected and modified.

Given the above, the claim that R is amongst the languages with the most
powerful \eval seems plausible. The rationale for R's design seems to have been
to expose as much of the language and its internals as possible in order to
maximize expressivity. In R, \eval is key tool to extend the language and
implement DSLs, it is also a replacement for macros. By contrast, the designers
of Julia chose to limit \eval. In Julia, only global variables can be
side-effected and environments cannot be readily manipulated. This is designed
to shield optimized code from some of the most pernicious uses of the
facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning
mechanism, called world age, to ensure that any methods defined within an \eval
only become visible at well-defined program points and thus that recompilation
does not have to occur when optimized code is running~\cite{oopsla20a}.

@@

The R language exposes a rich reflective interface with four main functions,
\c{eval}, \c{evalq}, \c{eval.parent}, and \c{local}. The first is the most
general with three arguments, an expression, its environment and enclosing
environment. The following simplifies details not relevant to this paper, the
interested reader should consult \citet{hadley}.

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
Parameter \c e is the \c{expression} to evaluate, \c{envir} is the evaluation
environment, and \c{encl} is used to look up variables not found in \c{envir}.
The default value of latter is either the caller's environment or top level.
\Eval can take several types for \c e, for our purposes on the \c{expression}.
These can be thought of as the abstract syntax trees returned by the parser. An
\c{expression} can be obtained by calling \c{quote} or \c{parse} wit a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Usually, expressions are obtained from promises using \c{substitute}. Consider a
function called with \c{x = a+b}, then \c f returns an expression object.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # expression(a+b)
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise bound to \c x. Function \c{substitute(e,envir)} takes two arguments, the
second is used as a substitution environment for free variables in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns the current environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Programmers sometimes use environments as hash maps as they are the
only data type with reference semantics and have a built-in string look up.
Functions \c{parent.frame} and \c{sys.frame} return environments further up the
call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
the argument to prevent it from being evaluated in the current environment. The
\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.

@@

\citet{ecoop11} provided the first large-scale study of the runtime behavior of
\eval in JavaScript. They dynamically analyzed a corpus of the 10,000 most
popular websites with an instrumented web browser to gather execution traces.
They show that \eval is pervasive with 82\% of the most popular websites using
it. The reasons for its use include the desire to load code on demand,
deserialization of JSON dataq and lightweight meta-programming to customize web
pages. While many uses were legitimate, just as many were unnecessary and could
be replaced with equivalent and safer code. They categorized inputs to \eval so
as to cover the vast majority of input strings. Restricting themselves to \eval
in which all named variables refer to the global scope, many patterns could be
replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
measure code coverage, so the numbers presented are a lower bound on the
possible behaviors. Furthermore, JavaScript usage in 2011 is likely different
from today, e.g. Node.js was not covered by Richards. More details about dynamic
analysis of JavaScript can be found in~\cite{liang}.

\citet{wang} analyzed use of dynamic features in 18 Python programs to find if
they affect file change-proneness. Files with dynamic features are significantly
more likely to be the subject of changes, than other files. Chen et al. looked
at the correlation between code changes and dynamic features, including \eval,
in 17 Python programs~\cite{chen}. They did not observe many uses of \eval.
\citet{oscar} performed an empirical study of the usage of dynamic features in
1,000 Smalltalk projects. While \eval itself is not present, Smalltalk has a
rich reflective interface. The authors found that reflective are used in less
than 2\% of methods. The most common reflective method is \c{perform:}; it send
a message that is specified by a string. These features are mostly used in the
core libraries.

\citet{bodden} looked at usage of reflection in the Java DaCapo benchmark suite.
They found that dynamic loading was triggered by the benchmark harness. The
harness then executes methods via reflection, this caused static analysis tools
to generate an incorrect call graph for the programs in DaCapo.

\citet{Arceri21} study \eval in JavaScript from a software security point of
view. The authors report that 53\% of the malware they studied used \eval as a
means to obfuscate attack code or to mount attacks. They propose an abstract
interpretation-based approach to analyzing dynamic languages. One must construct
a static approximation of the argument to \eval and then analyze possible
behaviors of the interpreter when evaluating the generated code.

\citet{ecoop12} had a short section on the usage of \eval in R. They found the
it widely used in R code with 8500 call sites in CRAN and 2.2 million dynamic
calls. The 15 most frequent call sites account for 88\% of those. The
\c{match.arg} function is the highest used one with 54\% of all calls. In the
other call sites, they saw two uses cases. The most common is the evaluation of
the source code of a promise retrieved by \c{substitute} in a new environment;
e.g. as done in the \c{with} function. The other use case is the invocation of a
function whose name or arguments are determined dynamically. For this purpose, R
provides \c{do.call} and thus \eval is overkill.
